Variational Monte Carlo on bosonic systems

### 2018-02-27 project_1/article/article.tex
Add drift force to the non-interacting system

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\ket{\Psi_T}$, we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_i^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

@@

    \subsection{The drift force}
        A disadvantage in the use of the brute-force Metropolis algorithm is
        that we might be spending much computational resources in an
        uninteresting part of configuration space. To make smarter moves we will
        use the Metropolis-Hastings algorithm (which will be discussed in due
        time). This algorithm is dependent on the drift force of the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \frac{2\nabla\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements.

        \subsubsection{Non-interacting harmonic oscillator}
            We start by finding an analytical expression for the local energy in
            a system where we set $a = 0$ and $\beta = 1$, i.e., a system of
            non-interacting harmonic oscillators.
            \begin{align}
                \Psi_T(\vf{r})
                = \Phi_T(\vf{r})
                = \prod_i^N \exp\bigl[
                    -\alpha |\vf{r}_i|^2
                \bigr],
            \end{align}
            where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
            $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
            (in the spherical case)
            \begin{align}
                H &= \sum_i^N h(\vf{r}_i)
                = \sum_i^N \Biggl(
                    -\frac{\hbar^2}{2m}\nabla_i^2
                    + \half m \omega^2 r_i^2
                \Biggr).
            \end{align}
            The hardest part of finding the local energy is computing the
            kinetic term. Starting with the gradient of the wavefunction in the
            non-interacting case, we get
            \begin{align}
                \nabla_k\Psi_T(\vf{r})
                &= -2\alpha \vf{r}_k\Psi_T(\vf{r}).
            \end{align}
            The Laplacian yields
            \begin{align}
                \nabla^2_k\Psi_T(\vf{r})
                &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
            \end{align}
            where $d$ is the dimensionality of the problem determined by
            $\vf{r}_k \in \mathbb{R}^d$. This yields the analytical expression
            for the local energy in the non-interacting case to be
            \begin{align}
                E_L(\vf{r})
                &=
                \sum_i^N
                \Biggl(
                    -\frac{\hbar^2}{2m}
                    \bigl[
                        -2d\alpha + 4\alpha^2 r_i^2
                    \bigr]
                    + \half m\omega^2 r_i^2
                \Biggr).
            \end{align}
            In natural units, with $\hbar = c = m = 1$, this reduces to
            \begin{align}
                E_L(\vf{r})
                &=
                \alpha dN
                + \biggl(
                    \half\omega^2
                    - 2\alpha^2
                \biggr)
                \sum_i^N r_i^2.
            \end{align}
            It is worth noting that for $\alpha = \half\omega$ we will find a
            stable value. This happens as the entire sum over all the random
            walkers disappears.

        \subsubsection{The full system}

@@

            \neq 0$ we write the trial wavefunction as

>>>>

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i = 1}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\ket{\Psi_T}$, we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_{i = 1}^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

@@

    \subsection{The drift force}
        A disadvantage in the use of the brute-force Metropolis algorithm is
        that we might be spending much computational resources in an
        uninteresting part of configuration space. To make smarter moves we will
        use the Metropolis-Hastings algorithm (which will be discussed in due
        time). This algorithm is dependent on the drift force of the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

\section{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$. We thus get the trial
    wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -2\alpha\vf{r}_k.
    \end{align}
    Using the Laplcian we can compute the kinteic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
    \end{align}
    It is worth noting that for $\alpha = \pm\half\omega$ we will find a stable
    value which turns out to be the exact energy minimum. This happens as the
    entire sum over all the random walkers disappears.

\section{Interacting hard sphere bosons}

@@

    \neq 0$ we can write the trial wavefunction as
